---
title: "Impacts of agroforestry practices on the biodiversity of Panamanian birds"
author: "Alex White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warnings=FALSE}
library(tidyverse)
library(ggplot2)
library(sf)
sf_use_s2(FALSE)
library(renv)
library(taxize)
```


## A game of names...

```{r}
coffee_birds <- read.csv("../data/coffee_Bird_Data.csv", stringsAsFactors = F)
head(coffee_birds)
```

We downloaded the ebird taxonomy from the Cornell Lab of Ornithology [here](https://www.birds.cornell.edu/clementschecklist/download/?__hstc=60209138.ef3edea8ce1040824b180bde7f920a82.1687545990397.1687545990397.1689258046438.2&__hssc=60209138.2.1689258046438&__hsfp=1390145134&_gl=1*1kufgl6*_ga*MTk2MzQwNTE1MC4xNjg3NDU5NDQ4*_ga_QR4NVXZ8BM*MTY4OTI1ODA0Ni4zLjEuMTY4OTI1ODA4NS4yMS4wLjA.&_ga=2.109099270.241330988.1689258046-1963405150.1687459448). The .csv is available in the `/data` folder in the github repo. 

```{r}
ebird_taxonomy <- read.csv("../data/ebird_taxonomy_v2022.csv", stringsAsFactors = F)
head(ebird_taxonomy)
```

```{r}
coffee_joined <- left_join(coffee_birds, ebird_taxonomy, by = join_by(Common_Name == PRIMARY_COM_NAME))
head(coffee_joined)
```


```{r}
coffee_species <- coffee_joined %>% filter(CATEGORY == "species")
```

We could also use a more generalized R package `taxize` to standardize the common names to latin species epithets ("genus_species"). We won't develop this method but the beginnings of this approach are shown below. 

```{r}
common_names <- coffee_birds %>% distinct(Common_Name)
common_names <- as.vector(common_names$Common_Name)
uid <- get_tsn(common_names[1:3], searchtype = "common", accepted = TRUE)
```

## Quantifying biodiversity

The goal of this exercise is to better understand how scientists quantify metrics of biodiversity and how those metrics are used to both make decisions in conservation and to understand fundamental processes of biology. Below we will work with the bird coffee dataset to quantify various metrics of diversity for different study plots in the coffee dataset, and to understand how values of diversity changes depending on what species features we select to represent biological diversity.  

### Taxonomic richness

Perhaps the simplest metric of biodiversity is taxonomic richness, or the number of individual taxonomic groups represented at the study location. Taxonomic groups can represent any taxonomic rank in the Linnaean taxonomic system (e.g., order, family, genus, species), and depending on the study, difference ranks are used as a unit of richness. All of the species in our dataset are birds, and so they are all in the same class *Aves*. The taxonomic rank in our dataset is mostly at the level of species, so it makes good sense to quantify and compare *species richness* across our sites. 

First we need to understand how many points are surveyed in our data. 

```{r}
point_names <- coffee_species %>% 
  distinct(Point_Name)
point_names <- as.vector(point_names$Point_Name)
length(point_names)
dim(coffee_species)
```

There are 11884 observations in 130 different locations in our dataset. Our goal is to count the number of unique species in each location. One *could* try to just count the unique number of observations per location. Does this reflect the quantity we are trying to estimate?

```{r}
species_richness <- coffee_species %>% count(Point_Name)
```

### For loops
Let's try something a bit more sophisticated. Digging deeper, it appears that these observations were collected on different days. The same species could be recorded multiple times per location. We need to approach the problem systematically. We can use a powerful tool called a `for` loop. 

* Loops are the fundamental structure for repetition in programming
* `for` loops perform the same action for each item in a list of things

```
for (item in list_of_items) {
  do_something(item)
}
```

In this case we want to build a `for` loop that loops over each location in our dataset and counts the number of unique species detected in that location. Following the template above we build our `for` loops as follows:

```{r}
for (place in point_names){ #start the for loop
  
  sp_rich <- coffee_species %>% # grab our original dataset
    filter(Point_Name == place) %>% # filter the dataset for each point name
    distinct(SCI_NAME) # select the distinct species names
  
  sp <- length(as.vector(sp_rich$SCI_NAME)) # generate a vector of the species names and provide the length of that vector
  
  #print(sp) # print the value of species richness
}
```

There is one obvious problem with the loop we've created - the results are merely printed. We need to store the results in an object (likely a vector), and we need to modify out loop in order to do that efficiently. 

### Looping with an index & storing results
R loops iterate over a series of values in a vector or other list like object. When we use that value directly this is called looping by value. But there is another way to loop, which is called looping by index. Looping by index loops over a list of integer index values, typically starting at 1. These integers are then used to access values in one or more vectors at the position indicated by the index.

We often use `i` to stand for “index” as the variable we update with each step through the loop.

If we modified our previous loop to use an index it would look like this:

```{r}
for (i in 1:length(point_names)){ #start the for loop
  
  sp_rich <- coffee_species %>% # grab our original dataset
    filter(Point_Name == i) %>% # filter the dataset for each point name
    distinct(SCI_NAME) # select the distinct species names
  
  sp <- length(as.vector(sp_rich$SCI_NAME)) # generate a vector of the species names and provide the length of that vector
  
  #print(sp) # print the value of species richness
}
```

Here we still need to add a method to store the results. To do this we start by creating an empty object the same length as the results will be before the loop starts. To store results in a vector we use the function `vector` to create an empty vector of the right length. `mode` is the type of data we are going to store. `length` is the length of the vector.

```{r}
sp <- vector(mode = "numeric", length = length(point_names))
```


```{r}
for (i in 1:length(point_names)){ #start the for loop
  
  sp_rich <- coffee_species %>% # grab our original dataset
    filter(Point_Name == point_names[i]) %>% # filter the dataset for each point name
    distinct(SCI_NAME) # select the distinct species names
  
  sp[i] <- length(as.vector(sp_rich$SCI_NAME)) # generate a vector of the species names and provide the length of that vector
  
}
sp  # print the value of species richness
```











# Appendix 

Exercises from https://datacarpentry.org/semester-biology/exercises/Loops-basic-for-loops-R/

## Basic For Loops (Loops)

Exercises from https://datacarpentry.org/semester-biology/exercises/Loops-basic-for-loops-R/

1. The code below prints the numbers 1 through 5 one line at a time. Modify it to print each of these numbers multiplied by 3.

```
numbers <- c(1, 2, 3, 4, 5)
for (number in numbers){
  print(number)
}
```

2. Write a for loop that loops over the following vector and prints out the mass in kilograms (mass_kg = 2.2 * mass_lb)

```
mass_lbs <- c(2.2, 3.5, 9.6, 1.2)
```

3. Complete the code below so that it prints out the name of each bird one line at a time.

```
birds = c('robin', 'woodpecker', 'blue jay', 'sparrow')
for (i in 1:length(_________)){
  print(birds[__])
}
```
4. Complete the code below so that it stores one area for each radius.

```
radius <- c(1.3, 2.1, 3.5)
areas <- vector(_____ = "numeric", length = ______)
for (__ in 1:length(________)){
  areas[__] <- pi * radius[i] ^ 2
}
areas
```

5. Complete the code below to calculate an area for each pair of lengths and widths, store the areas in a vector, and after they are all calculated print them out:

```
lengths = c(1.1, 2.2, 1.6)
widths = c(3.5, 2.4, 2.8)
areas <- vector(length = __________)
for (i in _____) {
  areas[__] <- lengths[__] * widths[__]
}
areas
```


